Effects of Solid State Drives on Database Performance

### 2017-05-01 term_paper/paper.tex
Paper might be finished?

    In the last decade or so, a new form of flash-based persistent storage, known as solid state drives (SSDs), have begun 
    to replace traditional mechanical hard disk drives (HDDs). They are orders of magnitude faster, lighter, more energy 
    efficient, and less prone to sudden impact than HDDs. As their prices fall and storage capacities increase, they have 
    taken more of a hold in both the consumer and enterprise market. These new SSDs behave differently than HDDs, which can 
    have a significant effect on any read or write heavy workload, such as database management systems. In this paper, we explore 
    what these effects are and their proposed solutions.

@@

For years, the gap in performance between processors and persistent storage has continually widened. Disk latency for 
traditional hard disk drives (HDDs) has improved by approximately ten percent per year, while we are just now beginning 
to see the end of Moore's Law for processors. \cite{Xie2011}

Traditional HDDs rely on mechanical parts, and therefore are destined to fail eventually, take relative eons to read 
or write data, and are prone to multiuple methods of failure, such as sudden force, vibration, or simply mechanical 
failure. They also are power inefficent and generate pleny of heat. \cite{Xie2011}

With the introduction of solid state drives (SSDs), we can begin to address many of the aforementioned issues, as they 
are not susceptible to the same problems. SSDs are lighter, faster, more energy efficient, and resistant to blunt force.
\cite{Xie2011}

However, they are not completely free from issue. At the time of writing, SSDs are still much more expensive per gigabyte 
than their HDD counterparts, though they promise to reach equivalence within a few years. They also have much different 
performance characteristics than HDDs, most notably, usually being worse at small random writes. They also tend to become 
slower as they fill up, and have a limited number of read/write/erase cycles they can endure before failing. 
\cite{Dumitru2007, Dirik2009, Xie2011}

Overall, SSDs are better than their predecessor, and will most likely come to replace them in the coming years. Therefore, 
applications that are heavily reliant on their storage medium, such as databases, must come prepared for it. In this paper, 
we will explore what effects these performance differences have as well as the solutions that have been proposed.

@@

Though solid state drives are designed to emulate the behavior of previous block devices, under the hood, they operate 
quite differently. \cite{Lee2008, Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016} In this section, we 
explain the general architecture of a solid state drive and explain these differences.

@@

The controller has many jobs. First, it must present to the host device the same interface HDDs do. Second, it must 
translate the commands issued to it to operations on the NAND flash. Third, it must wear the NAND in the drive evenly 
to prevent premature failure. Fourth, it must deal with the higher error rate that the NAND flash has over HDDs. 
\cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

To carry out these tasks, SSDs have a small computer, known as a microcontroller, embedded within them. The microcontroller 
is also paired with specialized hardware to speed up some of its computations, as well as some amount of DRAM to cache and 
buffer events. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

The controller must present to the host device the same interface other block devices do. However, the NAND flash is not 
accessed in the same way; instead, it has its own filesystem it maintains, known as the flash translation layer, or FTL. 
The FTL is used to perform the translation from the logical blocks the host requests to the pages and blocks within the 
NAND flash. \cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

Because cells must be erased before they can be written to, new data is always written to known empty pages, and the old 
pages are marked as garbage in the FTL. Therefore, the actual locations of data may move across the drive as it is written 
to, unlike HDDs. The FTL must maintain a list of empty, written, and garbage pages, and keep track of the mapping from 
the logical blocks requested by the host and their actual location on the SSD. The FTL also maintains the number of writes 
a block has experienced as well as a list of bad blocks that have either started out defective from the factory, or have 
worn out over time. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

When the controller is idle for some period of time, it will start performing garbage collection on the now invalid pages 
in order to free up more space. Some SSDs, especially earlier ones, would degrade in performance as the drive filled up 
and it ran out of free pages. The drive then had to immediately garbage collect to free up more, causing a delay in writes. 
Modern SSDs will understate their capacity to the host system, that way there are always free pages on the disk. 
\cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

NAND flash has a much higher error rate than the magnetic platters of HDDs. Therefore, all SSDs have some amount of 
hardware-assisted error correction. Higher quality drives have more dedicated hardware, while cheaper ones may share 
hardware for a larger number of pages. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

@@

Solid state drives have zero moving parts, hence their name. They do not require any spin up time, and their seek time is 
both uniform and much, much smaller. An additional benefit is that they consume much less energy. 
\cite{Cornwell2012, Micheloni2013, MatejFucek2014}

@@

Current systems tend to assume that they will be placing their data on to HDDs, and tend to optimize for that; they try 
to make up for the shortcomings of the physical media. They assume the location of the sectors requested are always in 
the same location, and that lower addresses are towards the faster, outside of the drive, and higher addresses are towards 
the inside. \cite{Cornwell2012}

Because of the seek time added by moving the head across the drive and the assumption of block locations, filesystem drivers 
will perform complicated math to determine how it can move the head the minimum distance for a sequence of reads and writes. 
They will also perform \textit{I/O coalescing,} or merging many smaller requests into one larger one, since HDDs are much 
better at large sequential operations. \cite{Cornwell2012}

@@

With the direction that computing is headed, data is invaluable. Data-heavy applications, such as those that heavily use 
online transaction processing (OLTP), require a data store that provides high performance and reliability.  The databases 
in which this data is stored are completely dependent on the storage medium in which they are backed. Its characteristics 
can seriously affect the end performance of the system. \cite{Lee2009, Xie2011, Micheloni2013, MatejFucek2014, Chen2016}

OLTP systems often require many small, random writes, which are the one thing that SSDs are terrible at. A small write 
requires rewriting the entire page with its changes to a separate location in the drive, then erasing the previous one, 
which is a costly operation. Many of these mall writes performed close together may lead to the SSD running out of free 
pages to write to, which would degrade performance tremendously. \cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

@@

SSD manufacturers have made numerous advantage on their end to help mitigate what they are bad at. There are several 
places apt for improvement, such as the FTL, or simply the amount of hardware included in the device. The more efficient 
the FTL is at translating and ordering requests and the more hardware that allows for increased parallelism, the more 
performance will come from the device. \cite{Xie2011, Chen2016}

@@

An improved interface for the device will also increase SSD performance. Currently, SSDs use the same interface that 
HDDs have employed, serial advanced technology attachment generation 3 (SATA-3), which is limited to somewhere around
600 MB/sec of thourughput. While that may be sufficent for HDDs, SSDs can easily outpace that. \cite{Micheloni2013}

Fortunately, there are faster interfaces present on modern day systems. The peripheral component interconnect express 
(PCIe) interface is used in all modern-day systems to connect important devices together. In fact, the SATA controller 
is often itself a PCIe device. PCIe promises transfer speeds anywhere from two to eight times that of SATA dependent on 
the number of lanes dedicated to the device. It also supports duplex and device-to-device communication, eliminating 
overhead. \cite{Micheloni2013}

@@

Systems that are data-heavy will need to attempt to optimize their access patterns for SSDs rather than HDDs.

>>>>

    In the last decade or so, a new form of flash-based persistent storage, known as solid state drives (SSDs), have begun 
    to replace traditional mechanical hard disk drives (HDDs). They are orders of magnitude faster, lighter, more energy 
    efficient, and less prone to sudden impact than HDDs. As their prices fall and storage capacities increase, they have 
    become much more popular in both the consumer and enterprise market. These new SSDs have different performance characteristics
    than mechanical drives, which can impact the performance on storage-heavy applications, such as databases. In this paper, 
    we explore what these impacts are and their proposed solutions.

@@

For years, the gap in performance between processors and persistent storage has continually widened. Disk latency for 
traditional hard disk drives (HDDs) has improved by approximately ten percent per year, but processor speed has nearly 
doubled each year according to Moore's Law until just recently. \cite{Xie2011}

Traditional HDDs rely on mechanical parts, which degrade over time, are prone to multiple methods of failure, such as 
sudden force, vibration, or simply mechanical failure. They also are inefficient at their jobs; they read and write data 
slowly while using lots of energy and generating lots of heat. \cite{Xie2011}

With the introduction of solid state drives (SSDs), they can address many of the aforementioned issues. SSDs are lighter, 
faster, more energy efficient, and resistant to blunt force. \cite{Xie2011}

However, they are not completely free from issue. Currently, SSDs are still much more expensive per gigabyte than their 
HDD counterparts, though they promise to reach equivalence within a few years. SSDs also have much different performance 
characteristics than HDDs. Most notably, SSDs are usually slower at small random writes. SSDs also tend to become slower 
as they fill up, and have a limited number of read/write/erase cycles they can endure before failing. 
\cite{Dumitru2007, Dirik2009, Xie2011}

Overall, SSDs are better than their predecessor, and will most likely replace HDDs in the coming years. The adoption of 
SSDs will impact applications that are heavily reliant on their storage medium, such as databases. In this paper, we will 
explore what effects these performance differences have as well as the solutions that have been proposed.

@@

Solid state drives are designed to emulate the behavior of previous block devices. However, they operate quite differently. 
\cite{Lee2008, Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016} In this section, we explain the general 
architecture of a solid state drive and explain these differences.

@@

The controller has many jobs. First, it must present to the host device the same interface HDDs do. Second, it must 
translate the commands issued to it to operations on the NAND flash. Third, it must wear the NAND in the drive evenly 
to prevent premature failure, which is known as wear leveling. Fourth, it must deal with the higher error rate that 
the NAND flash has over HDDs. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

To carry out these tasks, SSDs have a small computer, known as a microcontroller, embedded within them. The microcontroller 
is also paired with specialized hardware to speed up some of its computations, as well as some DRAM to cache and buffer 
events. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

The controller must present to the host device the same interface other block devices do. However, the NAND flash is not 
accessed in the same way; instead, it has its own filesystem it maintains, known as the flash translation layer (FTL). 
The FTL is used to perform the translation from the logical blocks the host requests to the pages and blocks within the 
NAND flash. \cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

Because cells must be erased before they can be written to, new data is always written to known empty pages, and the old 
pages are marked as garbage in the FTL. The actual locations of data may move across the drive as it is written to, unlike 
HDDs. The FTL must maintain a list of empty, written, and garbage pages, and keep track of the mapping from the logical 
blocks requested by the host and their actual location on the SSD. The FTL also maintains the number of writes a block 
has experienced as well as a list of bad blocks that have either started out defective from the factory, or have worn out 
over time. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

When the controller is idle for some period of time, it will start performing garbage collection on the marked invalid 
pages in order to free up more space. Some SSDs, especially earlier ones, would degrade in performance as the drive filled 
up and it ran out of free pages. The drive then had to immediately garbage collect to free up more, causing a delay in writes. 
Modern SSDs will understate their capacity to the host system, that way there are always free pages on the disk. 
\cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

NAND flash has a much higher error rate than the magnetic platters of HDDs. All SSDs have some amount of hardware-
assisted error correction to make up for this. Higher quality drives have more dedicated hardware, while cheaper ones 
may share hardware for a larger number of pages. \cite{Dirik2009, Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

@@

Solid state drives have zero moving parts; that is where the name solid-state comes from. They do not require any spin up 
time, and their seek time is both uniform and much, much smaller. An additional benefit is that they consume much less 
energy. \cite{Cornwell2012, Micheloni2013, MatejFucek2014}

@@

Current systems tend to assume that they will be placing their data on to HDDs, and tend to optimize for that; they try 
to make up for the shortcomings of the physical media. They assume the location of the sectors requested are always in 
the same location, and that lower addresses are towards the faster, outside of the drive, and higher addresses are towards 
the slower inside. \cite{Cornwell2012}

Because of the seek time added by moving the head across the drive and the assumption of block locations, filesystem drivers 
will compute complicated heuristics to determine how it can move the head the minimum distance for a sequence of reads 
and writes. They will also perform \textit{I/O coalescing,} or merging many smaller requests into one larger one, since 
HDDs are much better at large sequential operations. \cite{Cornwell2012}

@@

With the direction that computing is headed, data is invaluable. Data-heavy applications, such as those that heavily use 
online transaction processing (OLTP), require a data store that provides high performance and reliability.  The databases 
in which this data is stored are completely dependent on the storage medium that backs them. Its characteristics can 
greatly impact the end performance of the system. \cite{Lee2009, Xie2011, Micheloni2013, MatejFucek2014, Chen2016}

OLTP systems often require many small, random writes, which are the one thing that SSDs tend to perform worse at over 
HDDs. A small write requires rewriting the entire page with its changes to a separate location in the drive, then erasing 
the previous one, which is a costly operation. Many of these mall writes performed close together may lead to the SSD 
running out of free pages to write to, which would degrade performance tremendously. 
\cite{Cornwell2012, Micheloni2013, MatejFucek2014, Chen2016}

@@

SSD manufacturers have made numerous advances on their end to help mitigate what SSDs are bad at. There are several 
places apt for improvement, such as the FTL, or simply the amount of hardware included in the device. The more efficient 
the FTL is at translating and ordering requests and the more hardware that allows for increased parallelism, the more 
performance will come from the device. \cite{Xie2011, Chen2016}

@@

An improved interface for the device will also increase SSD performance. Currently, SSDs use the same interface that 
HDDs have employed, the Serial Advanced Technology Attachment generation 3 (SATA-3), which is limited to somewhere around
600 MB/sec of thourughput. While that may be sufficent for HDDs, SSDs can easily outpace that. \cite{Micheloni2013}

Fortunately, there are faster interfaces present on modern day systems. The Peripheral Component Interconnect Express 
(PCIe) interface is used in all modern-day systems to connect important devices together. In fact, the SATA controller 
is often itself a PCIe device. PCIe promises transfer speeds anywhere from two to eight times that of SATA dependent on 
the number of lanes dedicated to the device. It also supports duplex and device-to-device communication, eliminating 
overhead. \cite{Micheloni2013}

@@

Systems that are data-heavy will need to attempt to optimize their access patterns for SSDs rather than HDDs. Knowing that 
SSDs are slow at small, random writes, systems will want to avoid that behavior. Previous assumptions based on the workings 
of HDDs can be disregarded.
